Modeling MOOC Student Behavior With Two-Layer Hidden Markov Models

### 2017-02-22 short-paper/clickstream-hmm-short.tex
Fix typo in short paper's keywords.
Update short paper to use SIGCHI Proceedings format.

\documentclass{sigchi-ext}

% Please be sure that you have the dependencies (i.e., additional
% LaTeX packages) to compile this example.
\usepackage[T1]{fontenc}
\usepackage{textcomp}
\usepackage[scaled=.92]{helvet} % for proper fonts
\usepackage{graphicx} % for EPS use the graphics package instead
\usepackage{balance}  % for useful for balancing the last columns
\usepackage{booktabs} % for pretty table rules
\usepackage{ccicons}  % for Creative Commons citation icons
\usepackage{ragged2e} % for tighter hyphenation

% Some optional stuff you might like/need.
% \usepackage{marginnote}
% \usepackage[shortlabels]{enumitem}
% \usepackage{paralist}
\usepackage[utf8]{inputenc} % for a UTF8 editor only
\usepackage{amsmath}
\usepackage{float}
\usepackage[numbers]{natbib}

@@

\renewcommand\cite{\citep}
\def\meta/{\textsc{MeTA}}
\newcommand{\textretrieval}{textretrieval-001}
\newcommand{\sustain}{sustain-001}
\newcommand{\UIUC}{UIUC}

%% EXAMPLE BEGIN -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP --
% \copyrightinfo{Permission to make digital or hard copies of all or
% part of this work for personal or classroom use is granted without
% fee provided that copies are not made or distributed for profit or
% commercial advantage and that copies bear this notice and the full
% citation on the first page. Copyrights for components of this work
% owned by others than ACM must be honored. Abstracting with credit is
% permitted. To copy otherwise, or republish, to post on servers or to
% redistribute to lists, requires prior specific permission and/or a
% fee. Request permissions from permissions@acm.org.\\
% {\emph{CHI'14}}, April 26--May 1, 2014, Toronto, Canada. \\
% Copyright \copyright~2014 ACM ISBN/14/04...\$15.00. \\
% DOI string from ACM form confirmation}
%% EXAMPLE END

\copyrightinfo{Permission to make digital or hard copies of all or part of
  this work for personal or classroom use is granted without fee provided
  that copies are not made or distributed for profit or commercial
  advantage and that copies bear this notice and the full citation on the
  first page. Copyrights for components of this work owned by others than
  the author(s) must be honored. Abstracting with credit is permitted. To
  copy otherwise, or republish, to post on servers or to redistribute to
  lists, requires prior specific permission and/or a fee. Request
  permissions from Permissions@acm.org.\\
  \emph{L@S 2017}, April 20 - 21, 2017, Cambridge, MA, USA.\\
  Copyright is held by the owner/author(s). Publication rights licensed to
  ACM.\\
  ACM 978-1-4503-4450-0/17/04 \$15.00\\
  DOI: \url{http://dx.doi.org/10.1145/3051457.3053986}}

% Paper metadata (use plain text, for PDF inclusion and later
% re-using, if desired).  Use \emtpyauthor when submitting for review
% so you remain anonymous.
\def\plaintitle{Modeling MOOC Student Behavior with Two-Layer Hidden Markov
Models}
\def\plainauthor{Chase Geigle, ChengXiang Zhai}
\def\emptyauthor{}
\def\plainkeywords{MOOC log analysis; studetn behavior modeling; Markov models; hidden Markov models}
\def\plaingeneralterms{\plainkeywords}

\title{\plaintitle}

\numberofauthors{2}
% Notice how author names are alternately typesetted to appear ordered
% in 2-column format; i.e., the first 4 autors on the first column and
% the other 4 auhors on the second column. Actually, it's up to you to
% strictly adhere to this author notation.
\author{%
  \alignauthor{%
    \textbf{Chase Geigle}\\
    \affaddr{University of Illinois at Urbana-Champaign} \\
    \affaddr{201 N Goodwin Ave}\\
    \affaddr{Urbana, IL 61801} \\
    \email{geigle1@illinois.edu} } \vfil \alignauthor{%
    \textbf{ChengXiang Zhai}\\
    \affaddr{University of Illinois at Urbana-Champaign} \\
    \affaddr{201 N Goodwin Ave}\\
    \affaddr{Urbana, IL 61801} \\
    \email{czhai@illinois.edu} } }

% Make sure hyperref comes last of your loaded packages, to give it a
% fighting chance of not being over-written, since its job is to
% redefine many LaTeX commands.
\definecolor{linkColor}{RGB}{6,125,233}
\hypersetup{%
  pdftitle={\plaintitle},
%  pdfauthor={\plainauthor},
  pdfauthor={\emptyauthor},
  pdfkeywords={\plainkeywords},
  bookmarksnumbered,
  pdfstartview={FitH},
  colorlinks,
  citecolor=black,
  filecolor=black,
  linkcolor=black,
  urlcolor=linkColor,
  breaklinks=true,
}

% \reversemarginpar%

\newcommand{\citemeta}{\cite{Massung:2016:ACL}}

\begin{document}

\maketitle

% Uncomment to disable hyphenation (not recommended)
% https://twitter.com/anjirokhan/status/546046683331973120
\RaggedRight{}

% Do not change the page size or page settings.
\begin{abstract}
  Massive open online courses (MOOCs) provide educators with an abundance
  of data describing how students interact with the platform, but this data
  is highly underutilized today. This is in part due to the lack of
  sophisticated tools to provide interpretable and actionable summaries of
  huge amounts of MOOC activity present in log data. In this paper, we
  propose a method for automatically discovering student behavior patterns
  by leveraging the click log data that can be obtained from the MOOC
  platform itself in a completely unsupervised manner.
\end{abstract}

\keywords{\plainkeywords}

@@

\input{sec-intro}
\input{sec-model}
\input{sec-results}
\input{sec-conclusion}

\section{Acknowledgments}

This material is based upon work supported by the NSF GRFP under Grant
Number DGE-1144245.

\balance{}

\bibliographystyle{acm-sigchi-modern}
\bibliography{bib}

>>>>

%% SIGCHI Proceedings Format (modern) Sample Document
%% This sample file demonstrates the use of the `sigchi-modern' class file.
%%
%% The content of this document draws heavily on the `HCI Archive Format'
%% template supplied by SIGCHI, but has been modified to demonstrate and
%% document the LaTeX class.

% The parameter given to the `\documentclass' command can be one of:
%   `preprint', `submission', or `final'.
% These control the presence of certain features for each stage of the authoring
% process. For example, `submission' does not output the author block, and
% `final' disables page numbers.
%
% The \preprintonly{}, \submissiononly{}, and \finalonly{} commands allow you
% to wrap text that will only be output under the corresponding option.

\PassOptionsToPackage{inline}{enumitem}
\documentclass[final]{sigchi-modern}

\def\meta/{\preprintonly{\textsc{MeTA}}\finalonly{\textsc{MeTA}}%
\submissiononly{(redacted)}}

\newcommand{\citemeta}{\preprintonly{\cite{Massung:2016:ACL}}%
  \finalonly{\cite{Massung:2016:ACL}}%
  \submissiononly{[redacted citation]}}

% These packages aren't required, but do provide some helpful features that are
% used in this sample.
\usepackage{graphicx}  % Graphics for figures
\usepackage{booktabs}  % Nice formatting for tables
\usepackage{balance}   % Attempts to balance the columns on the last page

% A nice way to help avoid overful lines handing out into the margins is the
% `microtype' package, which lets certain characters extrude a *little* bit.
\usepackage{microtype}
% If the situation becomes dire, use:
%\sloppy

% You may use any bibliography package as long as the output matches that of
% the prescribed format. `natbib' is a fairly nice default, and the provided
% BibTeX style file has been tested with it.
\usepackage[numbers]{natbib}
\bibliographystyle{acm-sigchi-modern}

@@

% Define these tokens as appropriate for the conference to generate the
% necessary copyright notice for `final` mode.
\confname{L@S 2017}
\confdate{April 20-21}
\confyear{2017}
\conflocation{Cambridge, MA, USA}
\procissn{978-1-4503-4450-0/17/04}
\doi{10.1145/3051457.3053986}
% \copylicense be set to one of: \acmcopyright, \authorlicense, or \openlicense
%\copylicense{\authorlicense}

% End of preamble. Here comes the document.
\begin{document}

\title{Modeling MOOC Student Behavior With Two-Layer Hidden Markov Models}

% Author information can be set automatically if you provide a series of
% author names and affiliations:
%   \author[1]{Author One}
%   \author[2]{Author Two}
%   \author[1,2]{Author Three}
%   \affiliation[1]{Organisation One\\
%                   Some Country}
%   \affiliation[2]{Organisation Two\\
%                   Some Country}
%   \authorextra{\mailto{one@example.com}, \mailto{two@example.org},
%                \mailto{three@example.net}}
% Or, manually:
%   \author{
%     \authorname{Author One}\\
%     \authoraffil{
%       Organisation One\\
%       Some Country\\
%       \mailto{one@example.com}}}
%   \author{
%     \authorname{Author Two}\\
%     \authoraffil{
%       Organisation Two\\
%       Some Country\\
%       \mailto{two@example.org}}}
% Refer to the body text of the document for more details.
\author[1]{Chase Geigle}
\author[1]{ChengXiang Zhai}

% Set \authorwidth with the longest author name if you don't like the default
% spacing:
\settowidth{\authorwidth}{ChengXiang Zhai1}
\settowidth{\affilwidth}{University of Illinois at Urbana-ChampaignAAA}

\affiliation[1]{Department of Computer Science\\
    University of Illinois at Urbana-Champaign\\
    Urbana, Illinois, USA}

\authorpostscript{\email{\{%
  \href{mailto:geigle1@illinois.edu}{geigle1},
  \href{mailto:czhai@illinois.edu}{czhai}\}@illinois.edu}}

% If you format the author blocks manually, use \authorlist with a
% comma-delimited list of the author names.
% This is written into the PDF metadata for preprint and final modes.
%\authorlist{Author One, Author Two, Author Three, Author Four}

% Produces the title, author, and optional banner block.
\maketitle

\begin{abstract}
  Massive open online courses (MOOCs) provide educators with an abundance
  of data describing how students interact with the platform, but this data
  is highly underutilized today. This is in part due to the lack of
  sophisticated tools to provide interpretable and actionable summaries of
  huge amounts of MOOC activity present in log data. In this paper, we
  propose a new method for automatically discovering student behavior
  patterns by leveraging the click log data that can be obtained from the
  MOOC platform itself in a completely unsupervised manner.
\end{abstract}

% A semicolon-separated list of keywords to describe the paper.
\keywords{MOOC log analysis; student behavior modeling; Markov models;
hidden Markov models}

\newcommand{\acknowledgments}{This material is based upon work supported by
the National Science Foundation under Grant Number DGE-1144245 (GRFP) and
Grant Number 1629161.}

% ACM classification block.
% \category{}{}{}[] has three mandatory arguments and one optional argument; it
% can be used multiple times.
%\classification{%
%    \category{}{}{}
%}

@@

\input{../sec-intro}
\input{sec-model}
\input{sec-results}
\input{sec-conclusion}

\section{Acknowledgments}
\preprintonly{\acknowledgments{}}
\finalonly{\acknowledgments{}}
\submissiononly{(Anonymized for blind reviewing.)}

\balance
\bibliography{bib}

### 2017-02-22 short-paper/sec-conclusion.tex
Update short paper to use SIGCHI Proceedings format.

\section{Conclusions and Future Directions}
We proposed a two-layer hidden Markov model for MOOC student behavior
modeling on large collections of actions extracted from MOOC log data. This
model uniquely treats behavior patterns as a sequence of \emph{latent states},
rather than assigning these states in a rule-based manner, captures the
variable behaviors of students over time, and allows analysis at different
levels of granularity.

Although we only experimented with our model on one MOOC, the model is
completely general and can be easily applied to analyze the log of any
other course to enable deep understanding of student behaviors as well as
the correlations of such behaviors as other variables such as grades.
Our model can be used to produce an ordered list of latent states for each
student over time from which we could correlate course-wide drifts with
events in a course, which could for example allow the automatic discovery
of difficult segments of a course.

>>>>

\section{Conclusion}
% summarize the main contributions of the paper; summarize the experimental
% results; argue again for our novelty; discuss future directions
We proposed a two-layer hidden Markov model for MOOC student behavior
modeling in an unsupervised learning framework on large collections of
actions extracted from MOOC log data. This model is different from existing
methods in that it treats behavior patterns as a sequence of \emph{latent
states}, rather than assigning these states in a rule-based manner. It
captures the variable behaviors of students over time, and allows
analysis at different levels of granularity.

We showed that such a model does in fact capture meaningful behavior
patterns and produces descriptions of these behavior patterns that are easy
to interpret. We argued that it is important to capture student behavior
patterns with more sophisticated models than simple discrete distributions
over actions in order to capture information present in bigrams of actions
(or above). Finally, we investigated whether we can detect differences in
student behavior patterns as they correlate with course performance.
Specifically, we demonstrated that high-performing students produce
substantially different HMM transition diagrams that tend to show longer
concentration span in quiz-taking and more active forum participation as
compared with the average students.

\subsection{Future Work}

Although we only experimented with our model on one MOOC, the model is
completely general and can be easily applied to analyze the log of any
other course to enable deep understanding of student behaviors as well as
the coorrelations of such behaviors as other variables such as grades. We
plan to develop a MOOC log analysis system based on the proposed model to
both facilitate education research and help instructors improve course
design.

Specifically, our model can be used to produce an ordered list of latent
states for each student over time. Given these labeled sequences, we could
correlate course-wide drifts in these latent states with events in a
course. For example, we might be able to automatically discover difficult
or confusing parts of a course by noticing spikes in the distribution of
students over latent states over time.

%Our model does not explicitly model drop-out like
%\citet{Kizilcec:2013:LAK}, but doing so is an obvious extension. Our model
%should be able to provide predictions of when a student is ``at risk'' for
%dropping out under such an extension.
%
%Currently, the model learns a transition matrix over the latent states that
%is \emph{shared} across all students. It would be interesting to instead
%learn a different latent state transition matrix for each individual
%student, but keep the second-level Markov models shared. This would provide
%the model with more flexibility which may be desirable itself, but would
%also naturally result in a description of a student (via his or her HMM
%transitions) that could be incorporated into existing supervised learning
%techniques that try to predict student outcomes for understanding
%which of the latent behavior patterns discovered by TL-HMM are most predictive
%of the performance of student learning.

### 2017-02-22 short-paper/sec-model.tex
Update short paper to use SIGCHI Proceedings format.

\begin{margintable}
  \vspace{-4.7in}
  \begin{center}
    \begin{tabular}{rr}
      \multicolumn{2}{c}{\textbf{\textretrieval{}}}\\\hline
      Students & 18,941\\
      Sequences & 85,240\\
      Avg. Len. & 7.31
    \end{tabular}
    \caption{Statistics about the sequences extracted from the MOOC.}
    \label{table:datasets}
  \end{center}
\end{margintable}

\marginpar{%
  \vspace{-1.6in}
  Each sequence is comprised of the following actions:
  \begin{enumerate}[itemsep=0,align=left,leftmargin=*]
    \item quiz start,
    \item quiz submit,
    \item wiki (course material),
    \item forum list (view the list of all forums),
    \item forum thread list (view the list of all threads in a specific
      forum),
    \item forum thread view (view the list of posts within a specific
      thread),
    \item forum search (a search query issued against the forum),
    \item forum post thread (a new thread was posted),
    \item forum post reply (a new post was created within an existing
      thread), and
    \item view lecture (defined as either streaming or downloading a lecture
      video).
  \end{enumerate}
}

@@

An HMM is a specific probabilistic generative model with a ``built-in''
state transition system that would control the data to be generated by the
model, thus it is especially suitable for modeling sequence
data. We may say that students transition through different ``task states''
(or ``behavior states'')  in the process of study.  One such task state may
be to learn about a topic by mostly watching lecture videos, another task
state may be to work on quizzes, and yet another may be to participate in
forum discussions. While in each of these different states, the student
would tend to exhibit different surface ``micro'' behaviors.  For example,
in the lecture study state, the student would tend to have many
video-watching related behaviors and occasionally forum activities, while
in the quiz-taking state, the student would tend to show many quiz-related
``micro'' activities as well as asking questions or checking discussions on
the forum.

The most natural application of an HMM to analyze our data would treat each
latent state as having a discrete distribution over the surface ``micro''
behaviors, but this only gives a rather superficial characterization of the
student behavior. Instead, we want to directly characterize the transitions
between the ``micro'' actions, so we treat an {\em entire
sequence} of micro activities (e.g., one session of activities) as an
observed ``symbol'' from a latent state, and further model the generation
of such a sequence with another Markov model where we treat each micro
activity as an {\em observable} state, and model the transitions between
these activity states in very much the same way as the state transitions in
HMM. Adding this second layer would allow us to characterize a latent task
state in much more detail, as it would reveal not only what activities are
most common to a task state, but also the transition patterns between these
activities.

>>>>

@@

An HMM  is a specific probabilistic generative model with a ``built-in'' state transition system
that would control the data to be generated by the model, thus it is especially 
suitable for modeling sequence data~\cite{Rabiner:1990:RSR}. At any moment, the HMM would be in 
one of $k$ states $U=\{u_1, ...,u_k\}$, and at the next moment, the HMM would move to 
another state stochastically according to a transition matrix that specifies the probability of
going to state $u_i$ when the HMM is currently in state $u_j$, i.e., $p(u_i|u_j)$. 
When the HMM is in state $u$, the HMM can generate an observable data point $x$ 
according to an output probabilistic model $p(x|u)$. Thus if we ``run'' an HMM for 
$N$ time points denoted by $t=1, ..., N$, the HMM could ``generate'' a sequence of 
observations $x_1 ... x_N$, where each $x_i$ is an output symbol by 
going through a sequence of {\em hidden states} $w_1 ... w_N$ where $w_i \in U$ is a state. 
The association of such a latent sequence of state transitions with the observed symbols makes
it possible to use HMM to ``decode'' the latent behavior of students behind the surface behavior we directly observe in the log data, allowing for understanding student behavior more deeply than a model with no latent state variables.

In many ways, the generation process behind an HMM is meant to simulate the
actual behavior of a student. We assume that students transition through different ``task states'' (or ``behavior states'')  in the process of study. 
One such task state may be to learn about a topic by mostly watching
lecture videos, another task state may be to work on quizzes, and yet
another may be to participate in forum discussions. While in each of these
different states, the student would tend to exhibit different surface
``micro'' behaviors. For example, in the lecture study state, the student
would tend to have many video-watching related behaviors and occasionally
forum activities, while in the quiz-taking state (in order to pass each
module), the student would tend to show many quiz-related ``micro''
activities as well as asking questions or checking discussions on the
forum. Note that due to the complexity of the student behavior, it is very
difficult to accurately {\em prescribe} the specific surface ``micro''
behavior patterns for each state in advance, especially without  prior knowledge about the students. For example, forum activities are likely interleaved with other activities   in every task state and the interleaving pattern can be somewhat irregular with potentially many variations. 
The major motivations for using an HMM are that 
\begin{enumerate*}[label=(\arabic*)]
    \item it uses a probabilistic model (i.e., the output probability distribution $p(x|u)$ conditioned on each state) to directly capture the inevitable uncertainty in the association of surface ``micro'' activities with their corresponding latent task/behavior state, which is often our main target to discover and characterize, and 
    \item it does not make any assumption about which latent task/behavior
      state must be associated with which observed activities or how a
      student would move from one state to another, but instead allows our
      data to ``tell'' us what kind of associations are most likely,
      what kind of transitions are most probable, and which states tend to be more long-lasting for any particular set of students. 
\end{enumerate*}

However, if we use an ordinary HMM to analyze our data, we would treat each
observed ``micro'' activity (e.g., video watching, or forum post reading)
as an output symbol, and thus the output distribution $p(x|u)$ for each
discovered latent state would be a simple distribution over all kinds of
observable micro activities recorded in our log data (e.g., 50\% lecture
watching, 8\% quiz taking, 7\% quiz submission, 2\% course wiki reading,
...). While such a distribution is meaningful and can already help us
interpret the corresponding latent state, it only gives us a rather
superficial characterization of student behavior.

Ideally, we want $p(x|u)$ to characterize the directly observable ``micro''
behavior in more detail to further capture the relations and dependencies
of these micro activities. To this end, we would treat an {\em entire
sequence} of micro activities (e.g., one session of activities) as an
observed ``symbol'' from a latent state, and further model the generation
of such a sequence with another Markov model where we treat each micro
activity as an {\em observable} state, and model the transitions between
these activity states in very much the same way as the state transitions in
HMM.

Adding this second layer would allow us to characterize a latent task
state in much more detail, as it would reveal not only what activities are
most common to a task state, but also the transition patterns between these
``micro'' activities (e.g., it can reveal frequent back-and-forth
transitions between quiz-taking and quiz-submission, which would suggest a
concentrated period of taking quizzes). Combining this ``surface'' Markov
model over the ``micro'' actions with the ``deep'' hidden Markov model over
the latent task states gives us a fairly general and powerful two-layer HMM
(TL-HMM) that can simultaneously learn ``deeply'' the latent task/behavior
states and their transitions as well as the corresponding ``micro''
activity transition patterns associated with each latent state to
facilitate interpretation and analysis of the discovered latent state
patterns. Our implementation of the learning algorithm for TL-HMMs is
included as part of the \meta/ toolkit~\citemeta{}.

### 2017-02-22 short-paper/sec-results.tex
Update short paper to use SIGCHI Proceedings format.

\section{Experiment Results}

As an analysis tool, the TL-HMM model provides with us the following two
patterns to characterize student behavior:
\begin{enumerate*}[label=(\arabic*)]
  \item the latent state representations, and
  \item the latent state transitions.
\end{enumerate*}
Thus to evaluate the proposed model, we conduct experiments to
qualitatively analyze both types of patterns discovered from empirical MOOC
log data. Specifically, we look at the MOOC logs associated with the
\textretrieval{} Coursera MOOC offered by \UIUC{}.
Table~\ref{table:datasets} provides statistics about the sequences we
extracted from this MOOC and the action space we used for those sequences.

\begin{marginfigure}
  %\vspace{-2in}
  \centering
  \begin{subfigure}[t]{\textwidth}
    \includegraphics[width=0.95\textwidth,trim={0 3cm 0 2cm}]{../figures/trans-comp/trans-avg.png}
    \caption{\vspace{20pt}\label{fig:trans-avg}}
  \end{subfigure}

  \begin{subfigure}[t]{\textwidth}
    \includegraphics[width=0.95\textwidth,trim={0, 3cm 0 2cm}]{../figures/trans-comp/trans-perfect.png}
    \caption{\vspace{20pt}\label{fig:trans-perfect}}
  \end{subfigure}

  \begin{subfigure}[t]{\textwidth}
    \includegraphics[width=0.95\textwidth,trim={0, 3cm 0 2cm}]{../figures/trans-comp/trans-low.png}
    \caption{\label{fig:trans-low}}
  \end{subfigure}

  \caption{The latent state transition diagrams for a 4-state TL-HMM fit to
  \protect\textretrieval{} for all students (a) compared to only ``perfect''
  students (b) and only ``low'' students (c).}
  \label{fig:trans-comp}
\end{marginfigure}

@@

To visualize the Markov models that represent our latent states, we plot
them as a directed graph where we set the size of a node to be proportional
to its probability of being visited during a random walk.  We let the
thickness of a directed edge $(u, v)$ reflect the probability of taking
that edge given that a random walk is currently at note $u$ (as indicated
by the transition matrix). Figure~\ref{fig:states} shows the four latent
states we learned by fitting a 4-state TL-HMM to the \textretrieval{}
sequence dataset.

A unique property of our model is its ability to capture transitions
between the \emph{behavior patterns themselves} that are captured by the
latent states. In Figure~\ref{fig:trans-avg} we show the latent state
transition diagram for the 4-state TL-HMM we learned.
We can see that the ``forum browsing'' state 2 (figure~\ref{fig:state2})
has relatively lower probability than the other states as we might expect.
It also makes sense that state 0 (figure~\ref{fig:state0}) has rather high
probability as this state likely captures all sequences where a student
logged in to the platform and then did nothing else (likely checking for
updates). If we look at state 1 (figure~\ref{fig:state1}) and state 3
(figure~\ref{fig:state3}), their relative probabilities match our intuition
as well. State 1 seems to capture a more engaged browsing session, where
there is non-negligible probability associated with different activities
such as quiz taking and forum browsing and, importantly, these activities
have high probability symmetric edges (so students are taking quizzes one
after the other, or viewing forum threads in succession). By contrast,
state 3 seems to capture a more passive student, with negligible
probability mass associated with forum activity (with low symmetry in the
edges). The link between ``quiz submit'' and ``quiz start'' (indicating
quiz repetition) is also significantly lower than state 1.

Thus, we might expect to see students that perform well in the course
preferring states 1 and 2 over states 0 and 3. To verify this, we took the
model we learned on the full training data and retrofit it to training data
consisting only of sequences produced by students that had perfect marks.
To prevent the latent state meanings from drifting, we forced the model
parameters associated with their Markov model representations to be fixed,
in effect only learning initial and transition probabilities for the top
layer of our TL-HMM.\ We show the updated latent state transition diagram
in Figure~\ref{fig:trans-perfect}. We can see that the probability of state
2 has increased dramatically, while the probability of states 0 and 3 has
decreased. State 1 had its probability increase, but only very slightly.

In Figure~\ref{fig:trans-low} we plot the latent state transition diagram
for a second group of ``low'' students. These students were selected so
that they attempted all required quizzes in the course, but such that their
average quiz score was $\leq 70\%$. Here, we see that \emph{state 1} has a
large increase in size, where we might have expected state 3 to grow
instead.  However, there is an alternative explanation for this phenomenon.
Since state 1 seems to indicate a highly engaged student, it is a perfectly
reasonable explanation for the ``low'' student group as they are going to
be working hard to try to fill in the gaps in their knowledge. By contrast,
the ``perfect'' student group likely has many members who can take the quiz
more passively and get perfect marks, perhaps because they already know
much of the material being presented, or are just naturally strong and do
not require much background review to perform well. This also explains why
state 1 did not increase in size for the ``perfect'' group like we were
anticipating. Our model lends itself well to discovering this potentially
counter-intuitive insight.

>>>>

\newcommand{\textretrieval}{textretrieval-001}

\newcommand{\sustain}{sustain-001}

\newcommand{\UIUC}{UIUC}

\section{Experiment Results}
As an analysis tool, the TL-HMM model provides with us the following two
patterns to characterize student behavior:
\begin{enumerate*}[label=(\arabic*)]
  \item the latent state representations, and
  \item the latent state transitions.
\end{enumerate*}
Thus to evaluate the proposed model, we conduct experiments to
qualitatively analyze both types of patterns discovered from empirical MOOC
log data.

Specifically, we looked at the MOOC logs associated with the \textretrieval{}
Coursera MOOC offered by \UIUC{} and extracted a dataset consisting of
18,941 students who produced 85,240 sequences with an average length of
7.31. We used the following ten actions as our action set $\mathbf{A}$:
\begin{enumerate*}[label=(\arabic*)]
  \item quiz start,
  \item quiz submit,
  \item wiki (course material),
  \item forum list (view the list of all forums),
  \item forum thread list (view the list of all threads in a specific
    forum),
  \item forum thread view (view the list of posts within a specific
    thread),
  \item forum search (a search query issued against the forum),
  \item forum post thread (a new thread was posted),
  \item forum post reply (a new post was created within an existing
    thread), and
  \item view lecture (defined as either streaming or downloading a lecture
    video).
\end{enumerate*}

\subsection{Latent State Representations}
To visualize these Markov models that represent our latent states, we plot
them as a directed graph where we set the size of a node to be proportional
to its probability of being visited during a random walk.  We let the
thickness of a directed edge $(u, v)$ reflect the probability of taking
that edge given that a random walk is currently at note $u$ (as indicated
by the transition matrix)\footnote{We do not plot the transition
probabilities directly within the figure to ease readability; we instead
will mention relevant transition probabilities in the text as we discuss
the plots. The plots were created using python-igraph:
\url{http://igraph.org/python/}.}.

@@

Figure~\ref{fig:states} shows the latent state representations learned by
fitting a 4-state TL-HMM to the \textretrieval{} sequence dataset. Our
interpretations of the states is as follows:
\begin{enumerate*}[label=(\alph*)]
  \item \textbf{state 0} likely captures all sequences where a student
    logged in to the platform and did nothing else (likely just checking
    for updates);
  \item \textbf{state 1} seems to capture a more engaged browsing session,
    where there is non-negligible probability associated with different
    activities such as quiz taking and forum browsing and, importantly,
    these activities have high probability symmetric edges (so students are
    taking quizzes one after the other, or viewing forum threads in
    succession);
  \item \textbf{state 2} captures a ``forum browsing'' state, with most
    weight being placed on consecutive thread views; and
  \item \textbf{state 3} seems to capture a more passive student, with
    negligible probability mass associated with forum activity (with low
    symmetry in the edges). The link between ``quiz submit'' and ``quiz
    start'' (indicating quiz repetition) is also significantly lower than
    state 1.
\end{enumerate*}

\subsection{Transitions Between Latent States}
A unique property of our model is its ability to capture transitions
between the \emph{behavior patterns themselves} that are captured by the
latent states. In Figure~\ref{fig:trans-avg} we show the latent state
transition diagram for a 4-state TL-HMM fit on \textretrieval{}. We can
immediately observe two things: \begin{enumerate*}[label=(\arabic*)]
  \item each latent state has a very high ``staying'' probability, and
  \item the prevalence of each latent state matches our intuition.
\end{enumerate*}
In particular, we can see that the forum browsing state (state 2) has
relatively lower probability than the other states as we might expect. It
also makes sense that state 0 (low activity) has rather high probability.
If we look at state 1 and state 3, their relative probabilities match our
intuition as well: there should be more students exhibiting more passive
behaviors (state 3) than very active behaviors (state 1).

\begin{figure*}
  \vspace{12pt}
  \centering
  \begin{subfigure}[t]{0.25\textwidth}
    \includegraphics[width=\textwidth,trim={0 3cm 0 2cm}]{../figures/trans-comp/trans-avg.png}
    \caption{\label{fig:trans-avg}}
  \end{subfigure}%
  \begin{subfigure}[t]{0.25\textwidth}
    \includegraphics[width=\textwidth,trim={0, 3cm 0 2cm}]{../figures/trans-comp/trans-perfect.png}
    \caption{\label{fig:trans-perfect}}
  \end{subfigure}%
  \begin{subfigure}[t]{0.25\textwidth}
    \includegraphics[width=\textwidth,trim={0, 3cm 0 2cm}]{../figures/trans-comp/trans-low.png}
    \caption{\label{fig:trans-low}}
  \end{subfigure}
  \caption{The latent state transition diagrams for a 4-state TL-HMM fit to
  \protect\textretrieval{} for all students (a) compared to only ``perfect''
  students (b) and only ``low'' students (c).}
  \label{fig:trans-comp}
\end{figure*}

Thus, we might expect to see students that perform well in the course
preferring states 1 and 2 over states 0 and 3. To verify this, we took the
model we learned on the full training data and retrofit it to training data
consisting only of sequences produced by students in \textretrieval{} that
had perfect marks. To prevent the latent state meanings from drifting, we
forced the model parameters associated with their Markov model
representations to be fixed, in effect only learning initial and transition
probabilities for the top layer of our TL-HMM.\ We show the updated latent
state transition diagram in Figure~\ref{fig:trans-perfect}. We can clearly
see that the probability of state 2 has increased dramatically, consistent
with previous observations of the positive correlation between forum
activity and grades~\cite{Huang:2014:LAS}, while the probability of states
0 and 3 has decreased. State 1 had its probability increase very slightly.

In Figure~\ref{fig:trans-low} we plot the latent state transition diagram
for a second group of ``low'' students. These students were selected so
that they attempted all required quizzes in the course, but such that their
average quiz score was $\leq 70\%$. Here, we see that \emph{state 1} has a
large increase in size, where we might have expected state 3 to grow
instead.  However, there is an alternative explanation for this phenomenon.
Since state 1 seems to indicate a highly engaged student, it is a perfectly
reasonable explanation for the ``low'' student group as they are going to
be working hard to try to fill in the gaps in their knowledge. By contrast,
the ``perfect'' student group likely has many members who can take the quiz
more passively and get perfect marks, perhaps because they already know
much of the material being presented, or are just naturally strong and do
not require much background review to perform well. This also explains why
state 1 did not increase in size for the ``perfect'' group like we were
anticipating. \citet{Kizilcec:2017:CandE} observe similar phenomena with
the courses they studied where they find that certificate earning is
negatively correlated with help seeking behavior. Our model enables
data-driven discovery of potentially counter-intuitive insights like this.

%\begin{table}
%  \centering
%  \begin{tabular}{r|rrrrr}
%    \textbf{Group} & \textbf{State 0} & \textbf{State
%  1}\textsuperscript\textdagger &
%    \textbf{State 2}\textsuperscript\textdagger & \textbf{State 3} & \textbf{State 2
%    $\rightarrow$ 2}\textsuperscript\textdagger\\\hline
%    \textbf{Perfect} & 975.3  & 1001.5 & 999.0  & 1056.5 & 939.6\\
%    \textbf{Low}     & 1024.9 & 816.4  & 1230.5 & 1161.2 & 1187.4\\
%  \end{tabular}
%  \caption{Average rank for ``perfect'' and ``low'' student groups in the
%  ranked lists associated with the four latent states found by a TL-HMM.
%  $\dagger$ indicates statistically significant different mean ranks at $p
%  < 0.01$ according to an unpaired $t$-test.}
%  \label{table:mean-rank}
%\end{table}
%
%To quantify this finding, we perform the following experiment. First, we
%select all students from \textretrieval{} who completed all of the quizzes
%$\mathbf{L}_q \subset \mathbf{L}$. This gives us 1,985 students along with
%their average quiz score. We then create a ranked list of the students in
%$\mathbf{L}_q$ by sorting them by their ``preference'' for a specific
%latent state
%\begin{equation}
%  p_\ell(i) = \frac{\sum_{t=1}^T \gamma_t^{(\mathbf{o}_\ell)}(i)}
%  {\sum_{j=1}^{K} \sum_{t=1}^T \gamma_t^{(\mathbf{o}_\ell)}(j)}
%\end{equation}
%where $\mathbf{o}_\ell$ is the list of action sequences for student $\ell$
%and $\gamma$ is defined as before and computed using the Baum-Welch
%algorithm. We can now compare the average rank in this list for both
%the ``perfect'' student group and the ``low'' student group: a useful state
%for distinguishing the two groups should result in a ranked list with
%statistically significant differences in average rank between the two
%groups. Our results are summarized in Table~\ref{table:mean-rank}. Indeed,
%we discover that states 1 and 2 are correlated with the ``perfect'' or
%``low'' groups: state 1 ranks students in the ``low'' group significantly higher
%than those in the ``perfect'' group, and state 2 does the opposite and
%prefers students in the ``perfect'' group to those in the ``low'' group.
%
%Returning to Figure~\ref{fig:trans-comp}, we an also see a difference in
%the transitions between the latent states. In particular, look at the
%transitions in Figure~\ref{fig:trans-perfect} and
%Figure~\ref{fig:trans-low} that leave state 2. In the ``perfect'' group,
%nearly all of this probability mass is allocated for the self-loop ($p
%\approx 0.93$). In the ``low'' group, this self loop is less strong ($p
%\approx 0.80$; most easily seen by noting that the edges leaving state 2
%are darker than for the ``perfect'' group). We can perform a similar
%experiment to above by producing a ranked list of students $\ell \in
%\mathbf{L}_q$ by their staying probability for state 2 (that is, given that
%a student $\ell$ is already in state 2, how likely are they to remain there
%in the next action sequence?). This can be computed as
%\begin{equation}
%  p_\ell(2, 2) = \frac{\sum_{t=1}^{T-1} \xi_t^{(\mathbf{o}_\ell)}(2,2)}
%  {\sum_{i=1}^K \sum_{t=1}^{T-1} \xi_t^{(\mathbf{o}_\ell)}(2,i)}
%\end{equation}
%where $\xi$ is defined as before and computed using the Baum-Welch
%algorithm. The last column of Table~\ref{table:mean-rank} indicates that
%this transition feature also correlates with the achievement group and
%results in significant differences in mean rank.
